---
title: 'Mentored data science research with tweets'
author: "Frederick J. Boehm and Bret M. Hanlon"
date: "May 15, 2020"
params:
  lastmod: !r lubridate::now()
bibliography: jse.bib
geometry: "left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm"
output: 
  bookdown::pdf_book:
    toc: FALSE
    number_sections: TRUE
    includes:
      in_header: head.tex
---


```{r child = "abstract.Rmd"}
```

## Introduction

Twitter has changed the way that we communicate. 
In only 140 characters, users speak their piece on politics, 
current events, sports, media, and many other topics. 
Recent development of accessible statistical methods for text 
analysis now enable instructors to use tweets as pedagogical 
tools in mentoring undergraduate research projects. 

Some social media data, including tweets from Twitter, are 
available through website application product interfaces (APIs). Twitter 
shares, via a streaming API, a sample of approximately one percent of all 
tweets during an API query time period [@tweet_stream].
Researchers have studied tweets for a 
variety of purposes, including 
inference of relationships and social networks among users [@lin2011joint]; 
determination of 
authorship of specific tweets when multiple persons share a single account [@drob]; and 
study of rhetoric in recruiting political supporters [@pelled2018little;@wells2016trump].
Recognizing the potential utility of tweets for data science research and teaching, 
we created a 
collection of tweets over time by repeated querying of the Twitter streaming API.

@nolan2010computing argue for students to work with real 
data. 
Working with real data allows students to develop skill not 
only in statistical analysis,
but also in related data science skills, including data 
transfer from online sources, data storage, and using data 
from multiple file formats. 

Mentoring in the work place and in higher education can have 
many 
benefits, including improving students' development as thinkers and scholars, confidence in their 
own abilities, integration into the campus community,
and interest in graduate training 
[@baker2010beyond;@higgins2001reconceptualizing].

The content of tweets over time particularly intrigued us. We 
hypothesized that high-profile social media events would generate lots of tweets, and that we'd detect social media events through changes in tweet topic content over time.



We present below 1) an approach for collecting tweets in real 
time and 2) statistical methods for detecting social media 
events via topic modeling of tweets and 3) reflections on 
using this data set in research mentoring for undergraduate 
students.




## Methods

### Collecting tweets

We collected tweets by automated, repeated querying of Twitter's 
streaming API. This involved multiple steps. First, we created a 
new account on Twitter. With these user credentials, we used the 
R package `rtweet` to query the API. Because we work on computers 
with linux operating systems, we made use of linux software 
`crontab` to repeatedly execute R code to submit API queries. Each 
query lasts a user-specified duration. We timed the API queries so 
that there was no time lag between queries. We stored API query 
results in their native JSON format. The R package `rtweet` 
provides functions that parse tweet JSON to R data frames. We then 
conducted all further analyses with the R data frames.

As a proof of principle of our event detection strategy, we chose to 
examine tweets before, during, and after the National Football League's 
2015 Super Bowl. We fitted five latent Dirichlet allocation models. We examined five time periods, each lasting one hour in duration. The first period began approximately 48h before the Super Bowl halftime show. Subsequent time periods started at -24, 0, +24, and +48 hours after the approximate start time of the halftime show. 

We defined each time period to be a single corpus of tweets. We then fitted latent Dirichlet allocation models to each corpus. 








```{r child = "results.Rmd"}
```


```{r child = "discussion.Rmd"}
```








## References


