---
title: "Tweets"
author: "Frederick J. Boehm"
date: "4/9/2020"
output: html_document
---

## Goals

1. Write a R script to query twitter streaming api to get 1 percent sample over 5'.
1. Demonstrate how to set up a crontab to execute script every 5'.
1. Verify that crontab approach works by running it on my computer for, say, 15'.
    1. Cautions about twitter databases: deletions and amendments to tweets
1. Examine structure of JSON obtained from api. 
1. Create phony JSON to represent a collection of tweets.
    1. Use lorem ipsum for text of tweets

## R code to query twitter streaming api

Install `rtweet` package from CRAN.

```{r, eval = FALSE}
install.packages("rtweet")
install.packages("lubridate")
```

```{r}
rtweet::stream_tweets(
  timeout = (60 * 5),
  parse = TRUE,
  file_name = paste0("../data/", lubridate::now(), "-tweets")
)
```



### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies: 

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
